---
title: "Satellite Records of Flaring and Venting on Tribal Land"
author: "Jimmy Cloutier"
output: 
  html_document:
      code_folding: show
      df_print: paged
      toc: true
      toc_float: true
      code_download: true
---



```{r setup, message=FALSE, class.source="fold-hide"}

# For general data science
library(tidyverse)

# For reading Excel files
library(readxl)

# For data cleaning
library(janitor)

# For working with datetime
library(lubridate)

# For pretty tables
library(knitr)
library(gt)

knitr::opts_chunk$set(echo = TRUE)

```

# Introduction

The Howard Center checked operator-reported volumes from 2012 to 2020 against satellite recorded data [available](https://eogdata.mines.edu/download_global_flare.html) from the Earth Observation Group at the Colorado School of Mines. Also available from the website is a detailed [explanation](http://www.mdpi.com/1996-1073/9/1/14) by Christopher Elvidge of the Visible Infrared Imaging Radiometer Suite (VIIRS) technology used to measure flaring volumes from space. 

Users need to create an EOG account to download the data. 

# Load data

Using latitude-longitude coordinates the Howard Center plotted satellite-recorded flares to their corresponding U.S. state and saved it to new data sets available from the Howard Center [here](https://drive.google.com/file/d/1Kh13rdkcBjCkQHEM3V2NFjeVMkqQ-Qxc/view?usp=sharing). (For more information about how we spatially joined the data, please contact: howardcenter@asu.edu). 

```{r}

id <- "17-FBaY79bGB_ITt-73K4kIUnMHhN-pDS" # google file ID
flaring <- read.csv(sprintf("https://docs.google.com/uc?id=%s&export=download", id))


```


# Analysis

We filtered for the 13 states we investigated, then grouped by year and state to arrive at annual totals. These figures supports numerous facts in the project from total amounts flared to flaring trends over the period examined.

```{r}

flaring %>%
  filter(state_name == "Texas" | 
         state_name == "North Dakota" | 
         state_name == "New Mexico" | 
         state_name == "Alaska" | 
         state_name == "Wyoming" | 
         state_name == "Montana" | 
         state_name == "Utah" | 
         state_name == "Colorado" | 
         state_name == "Oklahoma" |
         state_name == "Pennsylvania" | 
         state_name == "West Virginia" |
         state_name == "Kansas" |
         state_name == "Louisiana" | 
         is.na(state_name)) %>%
  group_by(year, state_name) %>%
  summarise(cf = sum(cf, na.rm = TRUE),
            .groups = 'drop') %>%
  pivot_wider(names_from = year, values_from = cf) %>%
  mutate(`2020` = replace_na(`2020`,0), 
         "Total" = rowSums(across(c(`2012`:`2020`))),
         "Total (%)" = (Total/sum(Total))*100) %>%
  arrange(desc(Total)) %>%
  
# Pretty table
  gt( rowname_col = "state_name") %>% 
  tab_stubhead(label = "State") %>%
  tab_header(title = "Satellite Recorded Flaring",
             subtitle = "Flaring volumes in cubic feet recorded by VIIRS satellite from 2012 - 2020") %>%
  fmt_number(column = '2012':'Total (%)',
             decimals = 0) %>%
  grand_summary_rows(fns = list('Total' = "sum"), 
                           columns=('2012':'Total (%)'), 
                           formatter=gt::fmt_number, 
                           decimals=0)

```


FACT: According to satellite data, 81% of the flaring in Utah happens on tribal land, 17% on state land and 2% on federal land.

FACT: Satellite data analyzed by the Howard Center revealed that oil and gas operators on federal lands in New Mexico flared more than 138 billion cubic feet of gas, nearly half of the gas burned off in the state between 2012 and 2020.


```{r}

flaring %>%
  filter(admin != 'offshore') %>% # we don't need to look at offshore drilling for this, since it's not associated with any state
  group_by(state_name, admin) %>%
  summarise(cf = sum(cf, na.rm = TRUE),
            .groups = 'drop') %>%
  pivot_wider(names_from = admin, values_from = cf) %>%
  mutate(total=rowSums(across(tribal:state),na.rm = TRUE),
         pct_state = (state/(total)),
         pct_tribal = (tribal/(total)),
         pct_federal = (federal/(total))) %>%
  select(-state, -tribal, -federal) %>%
  arrange(desc(total)) %>%
  gt( rowname_col = "state" ) %>%
  fmt_number(column = total,
             decimals = 0) %>%
  fmt_percent(columns = pct_state:pct_federal, 
              decimals=0, 
              use_seps=TRUE) %>%
  cols_label(state_name = "State",
             total = "Total",
             pct_state = "% State",
             pct_tribal = "% Tribal", 
             pct_federal = "% Federal") %>%
  tab_header(title = "Satellite Recorded Flaring by Admin",
             subtitle = "Flaring volumes recorded by VIIRS satellite from 2012 - 2020")
  

```











