---
title: "Utah flaring and venting totals"
author: "Zoha Tunio and Sarah Hunt"
output:
  html_document: default
---

```{r setup, message=FALSE, class.source="fold-hide"}

# For general data science
library(tidyverse)

# For reading Excel files
library(readxl)

# For data cleaning
library(janitor)

# For working with datetime
library(lubridate)

# For pretty tables
library(knitr)
library(gt)

knitr::opts_chunk$set(echo = TRUE)

```

# Introduction

The Utah Division of Oil, Gas and Mining requires oil and gas operators on producing, shut-in and temporarily abandoned wells to fill a form titled `Form 11` (https://oilgas.ogm.utah.gov/pub/report_forms/frm_11_2005.pdf) with information on flared and vented gas. 

Data from the monthly reports (2012-present) are available at the division's website under its Data Research Center tab [here] (https://oilgas.ogm.utah.gov/oilgasweb/data-center/dc-main.xhtml). Click `Disposition Data (2009-2014; can open in Excel),` `Disposition Data (2015-2019; can open in Excel),` and `Disposition Data (2020-2024; can open in Excel)` to download the data sets for our selected time frame. This will download the data in .csv format. 


#Load Data

We compiled and cleaned the data from the state regulatory agency and saved it as a new dataset available from the Howard Center [here](https://drive.google.com/file/d/1R3FPAoFZDiJzdnMJpljVHrmak-TZu4dA/view?usp=sharing). (For more information about our data cleaning, please contact: howardcenter@asu.edu) 

We created a fresh csv file for the analysis: `ut_flaring_and_venting.csv`


```{r load_and_clean_data, echo=TRUE}

id <- "1R3FPAoFZDiJzdnMJpljVHrmak-TZu4dA" # google file ID
ut <- read.csv(sprintf("https://docs.google.com/uc?id=%s&export=download", id))

```


#Analysis

The story states: "Combined flared/vented gas volumes, as reported to the state: 11,390,778,000 cubic feet"

The story states: "According to state data, oil and gas companies in Utah flared and vented more than 11 billion cubic feet of natural gas between 2012 and 2020."

So, we're going to go ahead and summarize for the years 2012 to 2020. Remember that these results will be in MCF.


```{r}

ut %>%
  filter(year >= 2012 & year <= 2020) %>%
  group_by(year) %>%
  summarise(total = sum(Vol.VentedFlared, na.rm = TRUE)) %>%
  gt(rowname_col = "year") %>%
  grand_summary_rows(fns = list("total" = "sum")) %>%
  print()


```

